%==============================================================================
% ACKNOWLEDGEMENTS
%==============================================================================
\chapter*{Acknowledgements}
\addcontentsline{toc}{chapter}{Acknowledgements}
\markboth{Acknowledgements}{Acknowledgements}

\vspace{1cm}

I would like to express my deepest gratitude to my thesis supervisor for their continuous
guidance, invaluable insights, and unwavering encouragement throughout this research
journey. Their expertise in medical image analysis and deep learning has been
indispensable in shaping both the conceptual foundations and technical execution of
this work.

I am grateful to the organizers of the LDCTIQAC 2023 Grand Challenge at MICCAI 2023 for
providing the publicly available dataset of radiologist-annotated brain CT images, without
which this research would not have been possible. The detailed documentation by Lee et al.\
in their \textit{Medical Image Analysis} paper has been an essential reference throughout
this project.

I extend my thanks to the broader open-source community, particularly the developers of
PyTorch, the \textit{timm} library, and the scikit-learn project, whose tools formed the
computational backbone of this research.

I am also grateful to the faculty members of \textbf{\thesisschool}
whose teachings during my four years of undergraduate study provided
the theoretical and practical foundation upon which this project was built.

Finally, I am profoundly thankful to my family and friends for their patience and moral
support during the long hours of experimentation and writing. This work is dedicated to
everyone who has contributed, directly or indirectly, to my academic development at
Sichuan University.

\vspace{2cm}

\noindent
\textit{\thesisauthor}\\
\textit{\thesisschool, \thesismajor}\\
\textit{\thesisdate}
